\documentclass[twocolumn,a4paper]{article}
\usepackage[top=1cm, bottom=2.5cm, left=2.5cm, right=2.5cm]{geometry}
\usepackage{graphicx} % Required for inserting images
\usepackage{hyperref} % Required for \url command

\title{Simulation of a computer game by diffusion model\footnote{\url{https://github.com/MichaelPestuka/KNN_proj}}}
\author{
  Lukáš Foukal \and
  David Nepraš \and
  Michael Peštuka
}
\date{March 2026}

\begin{document}

\maketitle

% \textit{On 13.3. you have to submit 1 A4 in the system. Describe the problem you are solving, the approaches you draw inspiration from, data sets you are planning to use and a plan for your solution and experiments. You also have to include a link to a public GIT repository containing your project source codes. During the following week (16.-20.2.) you have to schedule a consultation with your project mentor to discuss the proposal.}

\section{Problem}
The goal of our project is to create a playable simulation of a computer game by using a diffusion model. For such a model, the input is the previous frame (or a sequence of frames) and the user action. The output is the next frame. Various computer games have already been simulated using this approach. We plan on simulating a retro platformer, e.g. Super Mario Brothers (1985).

Success on this type of task can be measured by the PSNR metric, for example.

\section{Data}
It is our intention to generate our own trainig data using pre-existing games. This data will be formatted as a sequence of frames and actions performed during those frames. To capture this data we can use a reinforcement learning platform such as Gymnasium \footnote{\url{https://openreview.net/pdf?id=qPMLvJxtPK}}, which will allow us to automatically generate a large amount of data.


\section{Approach and Experiments}

We have drawn inspiration from \textit{GameNGen}\footnote{\url{https://arxiv.org/pdf/2408.14837}} and \textit{Oasis}\footnote{\url{https://oasis-model.github.io/}}, simulating Minecraft and Doom respectively. Both comprise of an auto-encoder and a diffusion model. The former is based on pre-trained \textit{Stable Diffusion 1.4} models while the latter uses a transformer-based architecture with \textit{ViT}\footnote{\url{https://arxiv.org/pdf/2010.11929}} and \textit{DiT}\footnote{\url{https://arxiv.org/pdf/2212.09748}}.

In Appendix A of \textit{GameNGen}, the authors describe their inital experiment on the dinosaur game present in Google Chrome, noting that their approach worked well with limited training data and time. This shows us the feasibility of scaling down their approach based on pre-trained models.

\subsection{Experiments}

First experiments will involve finding the optimal architecture for the autoencoder itself. Since it is probable that the diffusion transformer will have multiple frames as input, making the encoded frames as small as possible will help with keeping the whole network lightweight. Experiments involving the diffusion transformer include trying different base models and experimenting with the amount of past frames to encode with the input.

\section{Work Plan}

The project will be carried out in several stages, starting with environment preparation and data collection, followed by model design, training, and evaluation.

\subsection{Game Selection and Environment Setup}

First, we will select a suitable game environment for the simulation task. The chosen game should have relatively simple visual dynamics and a limited action space so that the model can feasibly learn the transition dynamics. Potential candidates include retro platformers or simple arcade games available through the Gymnasium framework.

Once the game is selected, we will instrument the environment so that it provides access to rendered frames and user actions at each timestep. This will allow us to automatically collect gameplay trajectories suitable for training.

\subsection{Dataset Generation}

Training data will be generated by running automated gameplay sessions. These sessions will be produced either by simple heuristic agents or by random action policies to ensure a diverse set of game states. Each trajectory will consist of sequences of rendered frames and the corresponding actions.

% The dataset will then be processed into training samples where the input consists of a sequence of past frames and the action taken by the player, and the target output is the next frame. We will experiment with different sequence lengths to determine how much temporal context is necessary for stable prediction.

\subsection{Architecture Research and Design}

Next, we will investigate architectures suitable for conditional frame prediction. Inspired by prior work such as GameNGen and Oasis, we will explore a two-stage architecture consisting of:

\begin{itemize}
\item an autoencoder for compressing image frames into a latent representation,
\item a conditional generative model that predicts the next latent state based on previous frames and the player action.
\end{itemize}

Possible architectures for the generative component include diffusion transformers, convolutional diffusion models, and other conditional generative models capable of handling temporal inputs.

\subsection{Model Implementation and Initial Experiments}

After selecting promising architectures, we will implement prototype models and perform initial experiments. These experiments will focus on verifying that the model is capable of predicting plausible next frames and maintaining short-term temporal consistency.

% During this phase we will evaluate different architectural choices, including latent space dimensionality, number of past frames used as input, and different conditioning mechanisms for incorporating player actions.

\subsection{Training and Fine-tuning}

Once a suitable architecture is identified, we will train the model on the full dataset. Training will involve tuning hyperparameters such as learning rate, diffusion steps, and latent dimensionality. We will also experiment with fine-tuning pre-trained components where applicable, for example using pre-trained image encoders.

\subsection{Evaluation and Interactive Testing}

The final stage will involve evaluating the trained model in an interactive setting.
% The model will generate frames in real time based on the previous frames and user inputs, effectively simulating the game environment. We will analyze the visual quality, temporal stability, and responsiveness of the generated gameplay.

% Additionally, we will examine how the model behaves in longer rollouts, including potential artifacts or divergence from realistic gameplay dynamics.

\end{document}
